---
title: "dartR Introductory Tutorials"
author: "Arthur Georges and Emily Stringer"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: "cerulean"
    highlight: "tango"
    css: ./css/dartR_style.css
runtime: shiny_prerendered
description: "data in dartR"
---

<html lang="en">

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

#necessary to render tutorial correctly
library(learnr) 
library(htmltools)
#options(repos = BiocManager::repositories())
library(dartR.data)
library(dartR.base)
gl.set.verbosity(3)

library(glDemo) # for tutorial data

```

## Introduction

This is your hands on introduction to [dartR](https://github.com/green-striped-gecko/dartRverse?tab=readme-ov-file#dartrverse-). The worked examples and exercises in this tutorial correspond to the [[Intro to dartR ebook](http://dartr.biomatix.org/dartR)]{style="color:#ff0000;"}. There are also accompanying [AI podcast summaries](https://public.3.basecamp.com/p/Dp1F2egZGXY5siDf1MPFqyyg) for all the chapters, which you should definitely check out if you are more of an auditory learner.

If you can't see the left sidebar, you just need to make the tutorial panel wider.

### 

### 

Throughout this tutorial there will be interactive R scripts/consoles (like the one below). This is like a mini version of R running within this tutorial. But since we are already in R, you can also follow along in your own console and start writing your own code in the source panel. You could even try testing some of what you learn on your own data.

But first, try running the code below (press the button that says *run code*)

### R code box {data-progressive=FALSE}

```{r ex, exercise=TRUE}
# Press the Run Code button to run this code
1+1
```

```{r ex-solution}
# This is an example of the interactive component 
#of this tutorial. 

# Try running some code and testing out the three buttons:
#  - Run Code, 
#  - Start Over, 
#  - and Solution

```

###

Did you notice the tick that is showing before the *R cod box* heading after running the code?

### 

It is worth noting that each R code box is independent, so code you ran in one will not be recognised by another. Don't worry though, we have done some fancy coding in the background for you to continue along, keeping track of the tasks performed before hand. 

Any troubles or tribulations with coding in dartR, we have a great community that can be found on the [dartR google group](https://groups.google.com/g/dartr).

### 

Some of the exercises within this tutorial are specific to your working environment or Rstudio, these types of exercises will be surrounded by a green box, like the one below.

::: {.my-solution icon="false"}
It might be loading in data, or setting the working environment, or creating an R project.

They need to be run in your console. These are optional at the time, if you want you can continue with the worked example and come back to these when you want to start working in your own RStudio.
:::

### 

While going through the worked examples your progress will be saved, if at any point you would like to refresh the tutorial and start over, the <small>[Start Over]{style="color:#a3a3a3; font-family:'Jaldi', sans-serif;"}</small> button is located at the bottom of the left sidebar, below the tutorial content.

### 

One final note, don't forget that all aspects of the material are beneficial for learning. The ebook for being introduced to the theory and methods, the worked examples to gain experience in the application, and finally the exercises to apply everything you have learned.

Alright! Head back to the ebook and get ready to start working through the worked examples and exercises back here when prompted.

<p>Good luck on your Pop Gen journey! 😁</p>

## Worked Ex 1-1

yet to come...

## Exercises 1

## Worked Ex 2-1

yet to come...

## Exercises 2

## Worked Ex 3-1

yet to come...

## Exercises 3 

## Worked Ex. 4-1

[Basic Attributes and QC]{style="color:#cc9900; font-size: 16px; font-weight: bold;"}

### 

This worked example will take you by the hand and lead you through the analyses covered so far in Chapter 4. We will work with a small dataset to examine its class and interrogate its contents using [adegenet](https://adegenet.r-forge.r-project.org/) accessors.

### Examine dataset class

We will be using the `testset.gl` data for this worked example. First, copy the test dataset to a new genlight object called `gl`. You can then check the class of the object with `class()`.

```{r data1}
gl <- testset.gl

```

```{r ex1, exercise=TRUE}

```

```{r ex1-solution}
gl <- testset.gl
class(gl)
```

### 

The class is **dartR**, still a genlight object but with some additions.

### Examine dataset attributes

Examine the contents of your genlight object by simply typing its name. This will give you the attributes associated with the object.

```{r ex2, exercise=TRUE, exercise.setup = "data1"}

```

```{r ex2-solution}
gl
```

<small>Note: If it is a SNP dataset, the ploidy of each individual will be reported as (range 2-2). If it is a SilicoDArT dataset, the ploidy will be reported as (range: 1-1)</small>

### 

Displayed are the number of genotypes (individuals/specimens/samples), the size of the genlight object, and the number of missing values. The ploidy value should be 2-2 for SNP data for a diploid organism (dartR does not have support for polyploid organisms), so if it is something else, you have a problem with your data. SilicoDArT presence absence data has the ploidy set to 1-1.

### 

**Slots** containing important information are listed, such as `@position`, which lists the position of the SNPs in the sequence tags (referenced from 0 as position 1). The `@other` slot is particularly important, because it holds the loc.metrics from Diversity Arrays Technology Pty Ltd (DArT) and your ind.metrics.

### Compliance check

If any of the optional content slots indicated above are missing, consider running `gl.compliance.check()`. Try testing it with the current data. Make sure to save your compliant genlight object using `<-`.

```{r ex3, exercise=TRUE, exercise.setup = "data1"}

```

```{r ex3-solution}
gl <- gl.compliance.check(gl)
```

### 

This will render the genlight object compliant with dartR.

### Summary of the dataset

To obtain a basic summary for a genlight object, use `gl.report.basics()`.

```{r ex4, exercise=TRUE, exercise.setup = "data1"}

```

```{r ex4-solution}
gl.report.basics(gl)

```

### 

Note that **dartR** keeps a summary of manipulations applied to a genlight object, which can be useful if you wish to repeat the analysis or for documenting the steps taken in preparing the data for further analysis.

### Examine history

You can access the history of your dartR object with the following slot: `@other$history`. Try checking the history of `gl`.

```{r ex5, exercise=TRUE, exercise.setup = "data1"}

```

```{r ex5-solution}
gl@other$history

```

### 

There is not much to keep track of, yet! It might be worthwhile checking the history of `gl` as we go along to understand what is being tracked.  

### Examine metrics

To access metadata directly you can use commands of the form `gl@other$loc.metrics$CallRate`. Try saving the CallRate as a new r object named `cr` and then running a histogram using `hist()`. 

```{r ex6, exercise=TRUE, exercise.setup = "data1"}

```

```{r ex6-solution}
cr <- gl@other$loc.metrics$CallRate
hist(cr)
```

###

You can try this on other variables. To remind yourself of the variables in the metadata you can use `names()` for the `loc.metrics` and `ind.metrics`. 

```{r ex7, exercise=TRUE, exercise.setup = "data1"}

```

```{r ex7-solution}
names(gl@other$loc.metrics)
# or names(gl@other$ind.metrics)
```

###

You can also run `gl.report.basics`, which we ran previously.


### Using adegenet accessors

You can also interrogate the genlight object using commands built into the [adegenet](https://adegenet.r-forge.r-project.org/) package. Give each of these a try with `gl`.

`nLoc()` <br>
`locNames()`<br>
`nInd()`<br>
`indNames()`<br>
`nPop()`<br>
`popNames()`<br>
`pop()`

```{r ex8, exercise=TRUE, exercise.setup = "data1"}

```

### 

Note the distinction between `popNames(gl)` and `pop(gl)`. The two are related by pop names equals `unique(pop(gl))`.

### Convert to matrix

To convert your genlight object to a conventional matrix, use `as.matrix()`. Save your matrix as `m` and then using square brackets we can look at the first five individuals and seven loci `[1:5,1:7]`.

```{r ex9, exercise=TRUE, exercise.setup = "data1"}

```

```{r ex9-solution}
m <- as.matrix(gl)
m[1:5,1:7]
```

###

These are all useful for interrogating your genlight object, and of course can be used in your r scripts to subset and manipulate your data.


## Worked Ex. 4-2

Yet to come

## Exercises 4
